Exploring dimensions of biodiversity in Japanese ferns

### 2021-08-19 ms/manuscript.Rmd
Update MS
- Make figure part letters bold, small-caps, in parenthesis per J Biogeo style
- Add dependency on files used to build MS

```{r load-data}
# Load objects from targets workflow
tar_load(c(
  "repro_data",
  "occ_point_data_ferns",
  "occ_point_data_ferns_unfiltered",
  "biodiv_ferns_spatial",
  "k_taxonomy",
  "k_phylogeny",
  "signif_cells_protected_area",
  "protected_areas",
  "japan_shp",
  "phy_sig_results",
  "binary_sig_results",
  "traits_summary",
  "fern_traits",
  "japan_pterido_tree",
  "japan_fern_tree",
  "ppgi",
  "occ_point_data_ferns",
  "comm_ferns_full",
  "comm_ferns",
  "japan_points_raw",
  "plastome_calibration_dates",
  "plastome_tree",
  "green_list",
  "model_params",
  "lrt_results",
  "aic_env_repro",
  "model_fits",
  "biodiv_ferns_cent"
))
```

@@

Joel H. Nitta^1^, Brent D. Mishler^2^, Wataru Iwasaki^1,3,4,5,6^, and Atsushi Ebihara^7^

@@

^3^Department of Computational Biology and Medical Sciences, Graduate School of Frontier Sciences, The University of Tokyo, Kashiwa, Japan

^4^Atmosphere and Ocean Research Institute, The University of Tokyo, Kashiwa, Japan

^5^Institute for Quantitative Biosciences, The University of Tokyo, Tokyo, Japan

^6^Collaborative Research Institute for Innovative Microbiology, The University of Tokyo, Tokyo, Japan

^7^Department of Botany, National Museum of Nature and Science, Tsukuba, Japan

@@

@@

**Key words:** biodiversity, biogeography, phylogenetic diversity.

@@

Our data include all ferns known to occur within the political boundaries of Japan (20°N to 45°N, 122°E to 153°E). 
Japan consists of four main islands and thousands of smaller ones (`r figure("japan-map")`). 
The main islands have been in contact with the continent at various points in the past during periods of lower sea level, but the Ryukyu and Ogasawara archipelagos are oceanic islands that have never been connected to the mainland. 
The climate is subtropical in the south and temperate in other parts of the country. 
Map data were downloaded from the Geospatial Information Authority of Japan under the Creative Commons Attribution License v4.0 (https://www.gsi.go.jp/kankyochiri/gm_japan_e.html).

@@

We used a list of all non-hybrid fern specimens housed at herbaria in Japan. 
We chose to use this over other sources (e.g., GBIF), because of its high taxonomic quality: all identifications are standardized to a common taxonomy, the Japan Green List v. 1.01 (http://www.rdplants.org/gl/). 
Furthermore, the overall sampling completeness of this dataset is extremely high (`r s_figure("sampling-curve")`). 
The original list comprises `r nrow(occ_point_data_ferns_unfiltered) %>% scales::number(big.mark = ",")` specimens representing `r occ_point_data_ferns_unfiltered %>% pull(taxon) %>% n_distinct()` terminal taxa (hereafter defined as species, subspecies, forms, and varieties unless otherwise indicated). 
We filtered out any occurrences not within a map of standard grid squares defined by the Statistics Bureau of Japan (http://www.stat.go.jp/english/data/mesh/05.html).

@@

We used data compiled by @Ebihara2019b, which classifies each fern taxon as sexual, apomictic, mixed (both sexual and apomictic modes known from the same taxon), or unknown. 
We calculated the percentage of apomictic species in each grid cell as the sum of apomictic and mixed taxa divided by total number of taxa (`r s_figure("envir")`B).

@@

Sequences for the plastid, coding *rbcL* gene are available for ca. 97% of the Japanese fern flora [@Ebihara2019b]. 
However, using this dataset alone for phylogeny suffers two drawbacks: it is not densely sampled enough to resolve many internal nodes, and it lacks many lineages that have fossils available for molecular dating. 
Because one goal of the study is to understand the distribution of paleo- vs. neo-endemic areas (defined in units of time), we require an ultrametric tree. 
To obtain a maximally sampled, robust phylogenetic tree, we combined the *rbcL* data of @Ebihara2019b with a globally sampled plastid dataset including all fern sequences on GenBank for four plastid genes (*rbcL*, *atpB*, *aptA*, and *rps4*) and 66 other coding, single-copy plastid genes extracted from 100 complete fern plastomes (Nitta et al., in prep) (`r ape::Ntip(plastome_tree) %>% scales::number(big.mark = ",")` OTUs total). 
We conducted phylogenetic analysis with IQTREE [@Nguyen2015] and molecular dating with treePL [@Smith2012] using `r nrow(plastome_calibration_dates)` fossil calibration points after @Testo2016a. 
We then trimmed the tree to Japanese taxa only and used this for all further analyses.

@@

### Data accessibility

Code to replicate all analyses, figures, and this manuscript are available at https://github.com/joelnitta/japan_ferns_biogeo. 
A Docker image to run the code is available at https://hub.docker.com/repository/docker/joelnitta/japan_ferns_biogeo.
Data files, including the DNA alignment, phylogenetic tree, and community data matrix, are openly available from Figshare at http://doi.org/ADD_DOI_HERE_WHEN_ASSIGNED.

@@

The cleaned list of specimens included `r n_specimens` specimens representing `r n_specimens_taxa` taxa (`r n_specimens_taxa_percent`). 
The distribution of collection locations was highly skewed, with > 3,000 specimens per grid-cell northwest of Tokyo in Saitama prefecture, and many fewer elsewhere (`r s_figure("abundance")`A). 
Despite this, sampling redundancy was generally good [> 0.3; @Gonzalez-Orozco2014b] throughout the country (mean `r mean_redundancy`, final community data matrix; errors are SD unless mentioned otherwise) (`r s_figure("abundance")`B). 
The final community data matrix included `r ncol(comm_ferns)` taxa and `r nrow(comm_ferns) %>% scales::number(big.mark = ",")` grid-cells. `r n_excluded_cells` grid-cells were excluded due to low redundancy (< 0.1); these were mostly from Hokkaido and coastal areas.

@@

```{r phy-sig-stats}
# Verify that min lambda >0.95
assert_that(phy_sig_results %>% pull(lambda) %>% min > 0.95)
# Verify that max K < 0.0.6
assert_that(phy_sig_results %>% pull(kval) %>% max < 0.17)

@@

All continuous traits showed significant phylogenetic signal, but estimates of signal strength differed between *K* and lambda: all lambda values were > 0.95, indicating evolution by BM; however, values of *K* were < 0.17, indicating less phylogenetic signal than expected under BM (`r s_table("phy-sig")`). 
`r n_negative_d` out of `r nrow(binary_sig_results)` of continuous traits (coded as binary) showed as much or more phylogenetic signal as expected under BM (*D* =< 0); others had *D* > 0 (`r s_table("phy-sig-binary")`).

@@

Taxonomic richness is lowest on the northernmost island of Hokkaido, then increases with decreasing latitude until reaching a maximum in southern Kyushu, then decreases again in smaller islands further south (`r figure("raw-diversity")`A). 
Observed values of PD and FD were highly correlated with richness as expected (`r figure("raw-diversity")`B, C), showing an initial steep slope that gradually starts to level off at higher richness (`r s_figure("raw-biplot")`A, C). 
Observed values of RPD and RFD were not linearly related to richness, however, they showed high variance at low species richness, and low variance at high species richness (`r s_figure("raw-biplot")`B, D).

@@

Grid-cells with significantly low PD were almost all located on Honshu, the main island of Japan (`r figure("rand-diversity")`A). 
A smaller number of cells had significantly high PD, and these were located mostly on smaller southern (Ryukyu, Izu, Ogasawara) islands or coastal areas (Izu Peninsula, Kii Peninsula). 
The area near Yakushima includes cells with both significantly high and low PD. 
Grid-cells with significantly low RPD were mostly observed in southern Honshu, particularly in and around Kyushu (`r figure("rand-diversity")`B). 
Grid-cells with significantly high RPD are located both in the southern islands and northern Honshu and Hokkaido. 
Results of the randomization test for functional diversity were broadly similar to phylogenetic diversity, but with more significant cells for FD than PD and fewer significant cells for RFD than RPD (`r figure("rand-diversity")`C, D).

@@

All models were significant relative to null models without any fixed effects `r s_table("lrt-null")`. 
The effect of temperature dominated (had the greatest absolute *t*-value) in environmental models (`r figure("coeff")`A--E), and % apomictic taxa dominated in reproductive models (`r figure("coeff")`F--G); temperature and % apomictic taxa were significant in each model where they were included, whereas other predictors (area, precipitation, precipitation seasonality) were not significant in some models. 
Richness showed a hump-shaped relationship with temperature, reaching maximum richness at intermediate temperatures (`r figure("modelfit")`A). 
Phylogenetic and functional diversity showed either moderately (SES of PD, SES of FD) to strongly (SES of RPD, SES of RFD) hump-shaped relationships with temperature (`r figure("modelfit")`B--E). 
SES of RPD showed a clear, linear decline with % apomictic taxa (`r figure("modelfit")`G); a similar pattern was detected in SES of PD, but with wider confidence intervals (`r figure("modelfit")`F). 
Conditional AIC was lower (indicating better model fit) in the reproductive model compared to the environmental model for SES of RPD, but not SES of PD (`r s_table("aic")`).

@@

Analysis of bioregions grouped grid-cells into `r english2(k_taxonomy$optimal$k)` or `r english2(k_phylogeny$optimal$k)` clusters (`r s_figure("k-plot")`) on the basis of taxonomic or phylogenetic distances, respectively. 
The vast majority of grid-cells fell to a subset of these clusters: using taxonomic distances, they include Cluster 1 (Hokkaido and northern Honshu), Cluster 2 (southern Honshu), Cluster 3 (Ryukyu Islands, excluding Yakushima), and Cluster 4 (Ogasawara Islands) (`r figure("bioregions")`A; `r s_figure("dendrogram")`A). 
Phylogenetic distances produced similar results, but the Ryukyu and Ogasawara Islands merged into a single cluster (`r figure("bioregions")`B; `r s_figure("dendrogram")`B). 
The other much smaller clusters are likely artifacts of insufficiently low taxon richness for robust clustering, and are not discussed further.

@@

The latitudinal hump-shaped pattern of species richness observed in ferns here (`r figure("raw-diversity")`A) is well-documented [@Kusumoto2017; @Ebihara2019b], and similar patterns have been observed in deciduous broadleaved trees and conifers [@Shiono2015], and niche-modeled distributions of woody plants [@Fukaya2020]. 
Our spatial modeling analysis indicates that this is strongly linked to temperature, with maximum richness occurring at intermediate temperatures (`r figure("modelfit")`A).
Previous studies have identified temperature, water availability [@Bickford2006; @Qian2012], or a combination of these [@Nagalingum2015; @Link-Perez2018a; @Qian2021] to be important drivers of spatial fern diversity.
A hump-shaped pattern of richness along elevation (maximum richness at intermediate elevations) has been frequently observed in ferns on mountains in tropical or sub-tropical areas [@Kessler2011a]. 
Hypotheses to explain such a pattern include purely geometrical constraints as well as biological hypothesis positing maximally favorable environmental conditions at mid-elevations [@Colwell2004a]. 
Another reason for the greater richness in southern Honshu compared to smaller southern islands is that southern Honshu harbors high elevational variation that may support a wider range of taxa than is present in the southern islands.

The overwhelming pattern characterizing phylogenetic and functional diversity (PD and FD) is the high diversity of southern islands (Ryukyu and Ogasawara Isl.) (`r figure("rand-diversity")`A, C). 
This is almost certainly due to the presence of primarily tropical lineages that do not occur in other parts of the country; `r n_southern_genera` genera and `r n_southern_families` families only occur south of 30.1° latitude (just south of Yakushima Isl.). 
The similarity in patterns of PD and FD is likely due to the moderate degree of phylogenetic signal present in the traits used in our analysis, and suggests that in ferns, phylogeny can be used as a reasonable stand-in for functional diversity. 
Given their high phylogenetic diversity, it is perhaps not surprising that southern islands also host high amounts of phylogenetic endemism; indeed, the vast majority of cells with significant levels of PE, as well as paleoendemic cells, are located in the southern islands. 
This is clearly due to the small land area of these islands, which are the only areas with a subtropical climate in Japan.

Another striking pattern revealed here is the predominance of long branches in the northern half of Honshu and in southern islands, and the predominance of short branches in the southern half on Honshu, particularly Kyushu (`r figure("rand-diversity")`B). 
The long branches in the southern islands are likely being driven by the tropical lineages mentioned previously; those in the north may be due to refugial lineages that are distantly related to others and become more species rich at high-latitudes, such as *Equisetum* (`r s_figure("genus-map")`A). The preponderance of short branches in southern Honshu may be due to the radiation of certain species-rich genera in that area such as *Deparia* (`r s_figure("genus-map")`B) and *Dryopteris* (`r s_figure("genus-map")`D).

Furthermore, we identified % apomictic taxa as a strong driver of SES of RPD (`r figure("modelfit")`G).
Indeed, *Dryopteris* is one such genus with a high rate of apomictic taxa (`r s_figure("apo-genus")`), and other genera with high rates of apomixis (*Cyrtomium*, *Pteris*; `r s_figure("apo-genus")`) also reach greatest richness in southern Honshu (`r s_figure("genus-map")`).
This is consistent with the hypothesis that communities with high proportions of asexual taxa have lower genetic diversity, and is the first time to our knowledge that this has been demonstrated in ferns. 
We verified that this relationship is also observed when analyzed with a phylogeny with branch lengths in units of genetic change (`r ie`, the output of the maximum likelihood phylogenetic analysis, before dating), consistent with an effect on genetic diversity (`r s_figure("modelfit-not-ult")`B).
While temperature is also correlated with % apomictic taxa and therefore cannot be discounted, conditional AIC indicates a better fit of the reproductive model than the environmental model for SES of RPD (`r s_table("aic")`). 
Ultimately, it is likely that temperature, in combination with other environmental conditions, is driving the predominance of apomictic taxa: @Tanaka2014 also found that the proportion of apomictic taxa reaches a maximum at ca. 34.8 °N latitude and increases with temperature in Japan (their study was limited to Honshu and did not include the southern islands). 
@Tanaka2014 asserted that one limiting factor affecting establishment of apomictic ferns at high latitudes or cold climates is the large cell size of apomicts, which are typically polyploid. 
As fern gametophytes are only a single cell-layer in thickness with no cuticle or stomata, it is expected they would be particularly susceptible to freezing damage induced by larger cells with higher water content; however, no experimental studies have tested these hypotheses.

@@

The extreme evolutionary distinctiveness of southern islands is both a major feature of the Japanese fern flora and a challenge for analysis. 
The tropical lineages occurring in these islands are likely responsible for the comparatively low (`r ie`, clustered) values of PD observed on Honshu (`r figure("rand-diversity")`A). 
If a more restricted null model was used that did not consider all taxa capable of dispersing anywhere within the study area, different results would likely be obtained for the randomization test [@Mishler2020]. 
However, ferns are known to have high dispersal abilities [@Tryon1970; @Barrington1993; @Moran2001]. 
In our dataset, the taxon with the greatest latitudinal distribution, `r taxon_with_widest_lat$text`, spans nearly the entire region (`r taxon_with_widest_lat$min_lat`°N to `r taxon_with_widest_lat$max_lat`°N), and `r n_widespread_taxa` taxa have latitudinal distributions spanning the Tokara strait in the south (29°N) to the Tsugaru strait in the north (41.5°N). 
Therefore, it seems reasonable to assume that ferns are capable of dispersing across the entire study area.

@@

```{r raw-diversity-make}
# Richness
a <- ggplot() +
  geom_sf(data = japan, fill = "grey90", color = "black", lwd = 0.25) +
  geom_sf(data = biodiv_ferns_spatial, aes(fill = richness), color = "transparent", alpha = 0.8) +
  annotate(geom="richtext", y = Inf, x = -Inf, label = "<span style='color:black'>**A** Richness</span>", hjust = "inward", vjust = "inward", color = "white") + 
  labs(fill = "No. taxa") +
  scale_fill_scico(palette = "lajolla") + 
  scale_x_continuous(breaks = c(125, 135, 145)) +
  scale_y_continuous(breaks = c(25, 35, 45)) +
  map_theme()

# Raw PD
b <- ggplot() +
  geom_sf(data = japan, fill = "grey90", color = "black", lwd = 0.25) +
  geom_sf(data = biodiv_ferns_spatial, aes(fill = pd_obs), color = "transparent", alpha = 0.8) +
  annotate(geom="richtext", y = Inf, x = -Inf, label = "<span style='color:black'>**B** PD</span>", hjust = "inward", vjust = "inward", color = "white") +
  labs(fill = "Relative\nbranch\nlength") +
  scale_fill_scico(palette = "lajolla") + 
  scale_x_continuous(breaks = c(125, 135, 145)) +
  scale_y_continuous(breaks = c(25, 35, 45)) +
  map_theme() +
  theme(
    axis.text.y = element_blank()
  )

# Raw FD
c <- ggplot() +
  geom_sf(data = japan, fill = "grey90", color = "black", lwd = 0.25) +
  geom_sf(data = biodiv_ferns_spatial, aes(fill = fd_obs), color = "transparent", alpha = 0.8) +
  annotate(geom="richtext", y = Inf, x = -Inf, label = "<span style='color:black'>**C** FD</span>", hjust = "inward", vjust = "inward", color = "white") + 
  labs(fill = "Relative\nbranch\nlength") +
  scale_fill_scico(palette = "lajolla") + 
  scale_x_continuous(breaks = c(125, 135, 145)) +
  scale_y_continuous(breaks = c(25, 35, 45)) +
  map_theme() +
  theme(
    axis.text.y = element_blank()
  )

@@

**`r figure("raw-diversity")`**. Raw taxonomic (**A**), phylogenetic (**B**), and functional (**C**) diversity of the ferns of Japan. For **B** and **C**, raw branchlengths were transformed to relative values by dividing each branchlength by the sum of all branchlengths.

@@

```{r rand-diversity-make}
# PD
a <- ggplot() +
  geom_sf(data = japan, fill = "grey90", color = "black", lwd = 0.25) +
  geom_sf(data = biodiv_ferns_spatial, aes(fill = pd_signif), color = "transparent", alpha = 0.8) +
  annotate(geom="richtext", y = Inf, x = -Inf, label = "<span style='color:black'>**A** PD</span>", hjust = "inward", vjust = "inward", color = "white") + 
  scale_fill_manual(values = signif_cols) +
  map_theme() +
  theme(
    axis.text.x = element_blank(),
    legend.title = element_blank()
  )

>>>>

```{r load-data}
# Load objects from targets workflow
tar_load(c(
  "repro_data",
  "occ_point_data_ferns",
  "occ_point_data_ferns_unfiltered",
  "biodiv_ferns_spatial",
  "k_taxonomy",
  "k_phylogeny",
  "signif_cells_protected_area",
  "protected_areas",
  "japan_shp",
  "phy_sig_results",
  "binary_sig_results",
  "traits_summary",
  "fern_traits",
  "japan_pterido_tree",
  "japan_fern_tree",
  "ppgi",
  "occ_point_data_ferns",
  "comm_ferns_full",
  "comm_ferns",
  "japan_points_raw",
  "plastome_calibration_dates",
  "plastome_tree",
  "green_list",
  "model_params",
  "lrt_results",
  "aic_env_repro",
  "model_fits",
  "biodiv_ferns_cent",
  "refs_yaml",
  "refs_other_yaml",
  "template_file",
  "csl_file"
))
```

@@

Joel H. Nitta^1^, Brent D. Mishler^2^, Wataru Iwasaki^1,3^, and Atsushi Ebihara^4^

@@

^3^Department of Integrated Biosciences, Graduate School of Frontier Sciences, The University of Tokyo, Chiba, Japan

^4^Department of Botany, National Museum of Nature and Science, Tsukuba, Japan

@@

FIXME: add acknowledgements (not including reviewers)

@@

<!-- may include title words -->
**Key words:** apomixis, biodiversity, biogeography, ferns, Japan, phylogenetic diversity.

@@

Our data include all ferns known to occur within the boundaries of Japan (20°N to 45°N, 122°E to 153°E). 
Japan consists of four main islands and thousands of smaller ones (`r figure("japan-map")`). 
The main islands have been in contact with the continent at various points in the past during periods of lower sea level, but the Ryukyu and Ogasawara archipelagos are oceanic islands that have never been connected to the mainland. 
The climate is subtropical in the south and temperate in other parts of the country. 
Map data were downloaded from the Geospatial Information Authority of Japan under the Creative Commons Attribution License v4.0 (https://www.gsi.go.jp/kankyochiri/gm_japan_e.html).

@@

We used a list of all non-hybrid fern specimens housed at herbaria in Japan. 
We chose to use this over other sources (e.g., GBIF), because of its high taxonomic quality: all identifications are standardized to a common taxonomy, the Japan Green List v. 1.01 (http://www.rdplants.org/gl/). 
Furthermore, the overall sampling completeness of this dataset is extremely high (`r s_figure("sampling-curve")` in Appendix S1 in Supporting Information). 
The original list comprises `r nrow(occ_point_data_ferns_unfiltered) %>% scales::number(big.mark = ",")` specimens representing `r occ_point_data_ferns_unfiltered %>% pull(taxon) %>% n_distinct()` terminal taxa (hereafter defined as species, subspecies, forms, and varieties unless otherwise indicated). 
We filtered out any occurrences not within a map of standard grid squares defined by the Statistics Bureau of Japan (http://www.stat.go.jp/english/data/mesh/05.html).

@@

We used data compiled by @Ebihara2019b, which classifies each fern taxon as sexual, apomictic, mixed (both sexual and apomictic modes known from the same taxon), or unknown. 
We calculated the percentage of apomictic species in each grid cell as the sum of apomictic and mixed taxa divided by total number of taxa (`r s_figure("envir")`b).

@@

Sequences for the plastid, coding *rbcL* gene are available for ca. 97% of the Japanese fern flora [@Ebihara2019b]. 
However, using this dataset alone for phylogeny suffers two drawbacks: it is not densely sampled enough to resolve many internal nodes, and it lacks many lineages that have fossils available for molecular dating. 
Because one goal of the study is to understand the distribution of paleo- vs. neo-endemic areas (defined in units of time), we require an ultrametric tree. 
To obtain a maximally sampled, robust phylogenetic tree, we combined the *rbcL* data of @Ebihara2019b with a globally sampled plastid dataset including all fern sequences on GenBank for four plastid genes (*rbcL*, *atpB*, *aptA*, and *rps4*) and 66 other coding, single-copy plastid genes extracted from 100 complete fern plastomes (Nitta et al., in prep) (`r ape::Ntip(plastome_tree) %>% scales::number(big.mark = ",")` OTUs total). 
We conducted phylogenetic analysis with `r software("IQTREE")` [@Nguyen2015] and molecular dating with `r software("treePL")` [@Smith2012] using `r nrow(plastome_calibration_dates)` fossil calibration points after @Testo2016a. 
We then trimmed the tree to Japanese taxa only and used this for all further analyses.
Tree visualization was done with the `r pack("ggtree")` R package [@Yu2017].

@@

@@

The cleaned list of specimens included `r n_specimens` specimens representing `r n_specimens_taxa` taxa (`r n_specimens_taxa_percent`). 
The distribution of collection locations was highly skewed, with > 3,000 specimens per grid-cell northwest of Tokyo in Saitama prefecture, and many fewer elsewhere (`r s_figure("abundance")`a). 
Despite this, sampling redundancy was generally good [> 0.3; @Gonzalez-Orozco2014b] throughout the country (mean `r mean_redundancy`, final community data matrix; errors are SD unless mentioned otherwise) (`r s_figure("abundance")`b). 
The final community data matrix included `r ncol(comm_ferns)` taxa and `r nrow(comm_ferns) %>% scales::number(big.mark = ",")` grid-cells. `r n_excluded_cells` grid-cells were excluded due to low redundancy (< 0.1); these were mostly from Hokkaido and coastal areas.

@@

```{r phy-sig-stats}
# Verify that min lambda >0.95
assert_that(phy_sig_results %>% pull(lambda) %>% min > 0.95)
# Verify that max K < 0.6
assert_that(phy_sig_results %>% pull(kval) %>% max < 0.10)

@@

All continuous traits showed significant phylogenetic signal, but estimates of signal strength differed between *K* and lambda: all lambda values were > 0.95, indicating evolution by BM; however, values of *K* were < 0.1, indicating less phylogenetic signal than expected under BM (`r s_table("phy-sig")`). 
`r n_negative_d` out of `r nrow(binary_sig_results)` of continuous traits (coded as binary) showed as much or more phylogenetic signal as expected under BM (*D* =< 0); others had *D* > 0 (`r s_table("phy-sig-binary")`).

@@

Taxonomic richness is lowest on the northernmost island of Hokkaido, then increases with decreasing latitude until reaching a maximum in southern Kyushu, then decreases again in smaller islands further south (`r figure("raw-diversity")`a). 
Observed values of PD and FD were highly correlated with richness as expected (`r figure("raw-diversity")`b, c), showing an initial steep slope that gradually starts to level off at higher richness (`r s_figure("raw-biplot")`a, c). 
Observed values of RPD and RFD were not linearly related to richness, however, they showed high variance at low species richness, and low variance at high species richness (`r s_figure("raw-biplot")`b, d).

@@

Grid-cells with significantly low PD were almost all located on Honshu, the main island of Japan (`r figure("rand-diversity")`a). 
A smaller number of cells had significantly high PD, and these were located mostly on smaller southern (Ryukyu, Izu, Ogasawara) islands or coastal areas (Izu Peninsula, Kii Peninsula). 
The area near Yakushima includes cells with both significantly high and low PD. 
Grid-cells with significantly low RPD were mostly observed in southern Honshu, particularly in and around Kyushu (`r figure("rand-diversity")`b). 
Grid-cells with significantly high RPD are located both in the southern islands and northern Honshu and Hokkaido. 
Results of the randomization test for functional diversity were broadly similar to phylogenetic diversity, but with more significant cells for FD than PD and fewer significant cells for RFD than RPD (`r figure("rand-diversity")`c, d).

@@

All models were significant relative to null models without any fixed effects `r s_table("lrt-null")`. 
The effect of temperature dominated (had the greatest absolute *t*-value) in environmental models (`r figure("coeff")`a--e), and % apomictic taxa dominated in reproductive models (`r figure("coeff")`f--g); temperature and % apomictic taxa were significant in each model where they were included, whereas other predictors (area, precipitation, precipitation seasonality) were not significant in some models. 
Richness showed a hump-shaped relationship with temperature, reaching maximum richness at intermediate temperatures (`r figure("modelfit")`a). 
Phylogenetic and functional diversity showed either moderately (SES of PD, SES of FD) to strongly (SES of RPD, SES of RFD) hump-shaped relationships with temperature (`r figure("modelfit")`b--e). 
SES of RPD showed a clear, linear decline with % apomictic taxa (`r figure("modelfit")`g); a similar pattern was detected in SES of PD, but with wider confidence intervals (`r figure("modelfit")`f). 
Conditional AIC was lower (indicating better model fit) in the reproductive model compared to the environmental model for SES of RPD, but not SES of PD (`r s_table("aic")`).

@@

Analysis of bioregions grouped grid-cells into `r english2(k_taxonomy$optimal$k)` or `r english2(k_phylogeny$optimal$k)` clusters (`r s_figure("k-plot")`) on the basis of taxonomic or phylogenetic distances, respectively. 
The vast majority of grid-cells fell to a subset of these clusters: using taxonomic distances, they include Cluster 1 (Hokkaido and northern Honshu), Cluster 2 (southern Honshu), Cluster 3 (Ryukyu Islands, excluding Yakushima), and Cluster 4 (Ogasawara Islands) (`r figure("bioregions")`a; `r s_figure("dendrogram")`a). 
Phylogenetic distances produced similar results, but the Ryukyu and Ogasawara Islands merged into a single cluster (`r figure("bioregions")`b; `r s_figure("dendrogram")`b). 
The other much smaller clusters are likely artifacts of insufficiently low taxon richness for robust clustering, and are not discussed further.

@@

The latitudinal hump-shaped pattern of species richness observed in ferns here (`r figure("raw-diversity")`a) is well-documented [@Kusumoto2017; @Ebihara2019b], and similar patterns have been observed in deciduous broadleaved trees and conifers [@Shiono2015], and niche-modeled distributions of woody plants [@Fukaya2020]. 
Our spatial modeling analysis indicates that this is strongly linked to temperature, with maximum richness occurring at intermediate temperatures (`r figure("modelfit")`a).
Previous studies have identified temperature, water availability [@Bickford2006; @Qian2012], or a combination of these [@Nagalingum2015; @Link-Perez2018a; @Qian2021] to be important drivers of spatial fern diversity.
A hump-shaped pattern of richness along elevation (maximum richness at intermediate elevations) has been frequently observed in ferns on mountains in tropical or sub-tropical areas [@Kessler2011a]. 
Hypotheses to explain such a pattern include purely geometrical constraints as well as biological hypothesis positing maximally favorable environmental conditions at mid-elevations [@Colwell2004a]. 
Another reason for the greater richness in southern Honshu compared to smaller southern islands is that southern Honshu harbors high elevational variation that may support a wider range of taxa than is present in the southern islands.

The overwhelming pattern characterizing phylogenetic and functional diversity (PD and FD) is the high diversity of southern islands (Ryukyu and Ogasawara Isl.) (`r figure("rand-diversity")`a, c). 
This is almost certainly due to the presence of primarily tropical lineages that do not occur in other parts of the country; `r n_southern_genera` genera and `r n_southern_families` families only occur south of 30.1° latitude (just south of Yakushima Isl.). 
The similarity in patterns of PD and FD is likely due to the moderate degree of phylogenetic signal present in the traits used in our analysis, and suggests that in ferns, phylogeny can be used as a reasonable stand-in for functional diversity. 
Given their high phylogenetic diversity, it is perhaps not surprising that southern islands also host high amounts of phylogenetic endemism; indeed, the vast majority of cells with significant levels of PE, as well as paleoendemic cells, are located in the southern islands. 
This is clearly due to the small land area of these islands, which are the only areas with a subtropical climate in Japan.

Another striking pattern revealed here is the predominance of long branches in the northern half of Honshu and in southern islands, and the predominance of short branches in the southern half on Honshu, particularly Kyushu (`r figure("rand-diversity")`)b. 
The long branches in the southern islands are likely being driven by the tropical lineages mentioned previously; those in the north may be due to refugial lineages that are distantly related to others and become more species rich at high-latitudes, such as *Equisetum* (`r s_figure("genus-map")`a). The preponderance of short branches in southern Honshu may be due to the radiation of certain species-rich genera in that area such as *Deparia* (`r s_figure("genus-map")`b) and *Dryopteris* (`r s_figure("genus-map")`d).

Furthermore, we identified % apomictic taxa as a strong driver of SES of RPD (`r figure("modelfit")`g).
Indeed, *Dryopteris* is one such genus with a high rate of apomictic taxa (`r s_figure("apo-genus")`), and other genera with high rates of apomixis (*Cyrtomium*, *Pteris*; `r s_figure("apo-genus")`) also reach greatest richness in southern Honshu (`r s_figure("genus-map")`).
This is consistent with the hypothesis that communities with high proportions of asexual taxa have lower genetic diversity, and is the first time to our knowledge that this has been demonstrated in ferns. 
We verified that this relationship is also observed when analyzed with a phylogeny with branch lengths in units of genetic change (`r ie`, the output of the maximum likelihood phylogenetic analysis, before dating), consistent with an effect on genetic diversity (`r s_figure("modelfit-not-ult")`b).
While temperature is also correlated with % apomictic taxa and therefore cannot be discounted, conditional AIC indicates a better fit of the reproductive model than the environmental model for SES of RPD (`r s_table("aic")`). 
Ultimately, it is likely that temperature, in combination with other environmental conditions, is driving the predominance of apomictic taxa: @Tanaka2014 also found that the proportion of apomictic taxa reaches a maximum at ca. 34.8 °N latitude and increases with temperature in Japan (their study was limited to Honshu and did not include the southern islands). 
@Tanaka2014 asserted that one limiting factor affecting establishment of apomictic ferns at high latitudes or cold climates is the large cell size of apomicts, which are typically polyploid. 
As fern gametophytes are only a single cell-layer in thickness with no cuticle or stomata, it is expected they would be particularly susceptible to freezing damage induced by larger cells with higher water content; however, no experimental studies have tested these hypotheses.

@@

The extreme evolutionary distinctiveness of southern islands is both a major feature of the Japanese fern flora and a challenge for analysis. 
The tropical lineages occurring in these islands are likely responsible for the comparatively low (`r ie`, clustered) values of PD observed on Honshu (`r figure("rand-diversity")`a). 
If a more restricted null model was used that did not consider all taxa capable of dispersing anywhere within the study area, different results would likely be obtained for the randomization test [@Mishler2020]. 
However, ferns are known to have high dispersal abilities [@Tryon1970; @Barrington1993; @Moran2001]. 
In our dataset, the taxon with the greatest latitudinal distribution, `r taxon_with_widest_lat$text`, spans nearly the entire region (`r taxon_with_widest_lat$min_lat`°N to `r taxon_with_widest_lat$max_lat`°N), and `r n_widespread_taxa` taxa have latitudinal distributions spanning the Tokara strait in the south (29°N) to the Tsugaru strait in the north (41.5°N). 
Therefore, it seems reasonable to assume that ferns are capable of dispersing across the entire study area.

@@

```{r raw-diversity-make}
# Richness
a <- ggplot() +
  geom_sf(data = japan, fill = "grey90", color = "black", lwd = 0.25) +
  geom_sf(data = biodiv_ferns_spatial, aes(fill = richness), color = "transparent", alpha = 0.8) +
  annotate(geom="richtext", y = Inf, x = -Inf, label = "<span style='color:black'>**(a)** Richness</span>", hjust = "inward", vjust = "inward", color = "white") + 
  labs(fill = "No. taxa") +
  scale_fill_scico(palette = "lajolla") + 
  scale_x_continuous(breaks = c(125, 135, 145)) +
  scale_y_continuous(breaks = c(25, 35, 45)) +
  map_theme()

# Raw PD
b <- ggplot() +
  geom_sf(data = japan, fill = "grey90", color = "black", lwd = 0.25) +
  geom_sf(data = biodiv_ferns_spatial, aes(fill = pd_obs), color = "transparent", alpha = 0.8) +
  annotate(geom="richtext", y = Inf, x = -Inf, label = "<span style='color:black'>**(b)** PD</span>", hjust = "inward", vjust = "inward", color = "white") +
  labs(fill = "Relative\nbranch\nlength") +
  scale_fill_scico(palette = "lajolla") + 
  scale_x_continuous(breaks = c(125, 135, 145)) +
  scale_y_continuous(breaks = c(25, 35, 45)) +
  map_theme() +
  theme(
    axis.text.y = element_blank()
  )

# Raw FD
c <- ggplot() +
  geom_sf(data = japan, fill = "grey90", color = "black", lwd = 0.25) +
  geom_sf(data = biodiv_ferns_spatial, aes(fill = fd_obs), color = "transparent", alpha = 0.8) +
  annotate(geom="richtext", y = Inf, x = -Inf, label = "<span style='color:black'>**(c)** FD</span>", hjust = "inward", vjust = "inward", color = "white") + 
  labs(fill = "Relative\nbranch\nlength") +
  scale_fill_scico(palette = "lajolla") + 
  scale_x_continuous(breaks = c(125, 135, 145)) +
  scale_y_continuous(breaks = c(25, 35, 45)) +
  map_theme() +
  theme(
    axis.text.y = element_blank()
  )

@@

**`r figure("raw-diversity")`**. Raw taxonomic (a), phylogenetic (b), and functional (c) diversity of the ferns of Japan. For (b) and (c), raw branchlengths were transformed to relative values by dividing each branchlength by the sum of all branchlengths.

@@

```{r rand-diversity-make}
# PD
a <- ggplot() +
  geom_sf(data = japan, fill = "grey90", color = "black", lwd = 0.25) +
  geom_sf(data = biodiv_ferns_spatial, aes(fill = pd_signif), color = "transparent", alpha = 0.8) +
  annotate(geom="richtext", y = Inf, x = -Inf, label = "<span style='color:black'>**(a)** PD</span>", hjust = "inward", vjust = "inward", color = "white") + 
  scale_fill_manual(values = signif_cols) +
  map_theme() +
  theme(
    axis.text.x = element_blank(),
    legend.title = element_blank()
  )
